---
title: "Toothgrowth Analysis"
author: "Dominic Lloyd"
date: "October 24, 2015"
output: pdf_document
---


# Overview
Now we're going to analyze the ToothGrowth data in the R datasets package.  
 - Load the ToothGrowth data and perform some basic exploratory data analyses  
 - Provide a basic summary of the data.  
 - Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)  
 - State your conclusions and the assumptions needed for your conclusions.  
  
Some criteria that you will be evaluated on  
 - Did you  perform an exploratory data analysis of at least a single plot or table highlighting basic features of the data?  
 - Did the student perform some relevant confidence intervals and/or tests?  
 - Were the results of the tests and/or intervals interpreted in the context of the problem correctly?  
 - Did the student describe the assumptions needed for their conclusions?  
  
# Exploratory Data Analysis
```{r}
rm(list=ls())
data("ToothGrowth")
```
Output of ?ToothGrowth
The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).  
The dataset shows The Effect of Vitamin C on Tooth Growth in Guinea Pigs  
The dataset is a data frame with 60 observations on 3 variables.  
[,1]   len	 numeric	 Tooth length  
[,2]	 supp	 factor	 Supplement type (VC or OJ).  
[,3]	 dose	 numeric	 Dose in milligrams.  
```{r}
head(ToothGrowth)
str(ToothGrowth)
```


